---
title: "Computing insulin surrogate indices with SurInsulot"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Computing insulin surrogate indices with SurInsulot}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(SurInsulot)
```

## Overview

SurInsulot turns the surrogate indices of insulin sensitivity and secretion
into small, tested functions. There are two ways to use it:

1. **Call an index directly** when you want one number and full control over
   units.
2. **Call `compute_surrogates()`** on a data frame to derive the whole panel at
   once.

## Units: the one thing to get right

Most disagreements between implementations come down to units. Each function's
help page states exactly what it expects. The recurring conventions are:

* insulin in **uU/mL** (= mU/L); a few indices (the Stumvoll indices, OGIS) want
  **pmol/L** and say so,
* glucose in **mmol/L**, except the mg/dL-based indices (QUICKI, TyG, METS-IR,
  SPISE, Matsuda, Gutt, CIR, Kadowaki),
* C-peptide in nmol/L, glucagon and proinsulin in pmol/L.

Converters are provided:

```{r}
mgdl_to_mmol_glucose(90)
insulin_uU_to_pmol(10)
```

With `compute_surrogates()` you don't convert anything yourself: supply your data
in whatever units you have and declare them with `glucose_unit`, `insulin_unit`,
`cpeptide_unit`, `trig_unit`, `hdl_unit`, and `glucagon_unit` (the defaults are
mg/dL glucose and uU/mL insulin). Inputs are normalised before any index is
computed, so the results are identical regardless of the units you start from.

## Fasting indices

```{r}
homa_ir(insulin = 12, glucose = 5.56)   # insulin uU/mL, glucose mmol/L
quicki(insulin = 12, glucose = 100)     # QUICKI uses glucose in mg/dL
1 / homa_ir(insulin = 12, glucose = 5.56)  # reciprocal-HOMA sensitivity
```

## From an OGTT curve

The insulinogenic index and the Matsuda index summarise the OGTT. `igi()` works
at any timepoint; `matsuda_from_curve()` computes the required means for you.

```{r}
times <- c(0, 30, 60, 90, 120)
glucose <- c(95, 160, 150, 130, 110)   # mg/dL
insulin <- c(10, 80, 90, 70, 45)       # uU/mL

igi(insulin_0 = insulin[1], insulin_t = insulin[2],
    glucose_0 = 95 / 18, glucose_t = 160 / 18)   # 0-30 min, glucose mmol/L

matsuda_from_curve(times, glucose, insulin)             # full curve
matsuda_from_curve(times, glucose, insulin, upto = 30)  # modified 0-30
```

Trapezoidal areas are available on their own:

```{r}
auc_trapezoid(times, glucose)     # total AUC
iauc_trapezoid(times, glucose)    # incremental (baseline-subtracted) AUC
```

## The whole panel at once

`compute_surrogates()` takes a data frame using the GRADE/DPP short column names
and appends one column per estimable index. Missing inputs simply switch off the
indices that need them.

```{r}
df <- data.frame(
  g000 = c(95, 110), g030 = c(160, 190), g120 = c(140, 170),
  i000 = c(10, 18),  i030 = c(80, 60),
  trig = c(150, 200), chdl = c(45, 38), bmi = c(29, 34), sex = c(1, 2)
)

out <- compute_surrogates(df, verbose = FALSE)
round(out[, c("homa_ir", "quicki", "tyg", "matsuda_30", "igi_30")], 3)
```

The names of the computed indices are recorded as an attribute:

```{r}
attr(out, "surinsulot_computed")
```

## Disposition indices

A disposition index is secretion scaled by sensitivity — just their product.
Pass a secretion surrogate and a *sensitivity* surrogate (higher = more
sensitive):

```{r}
ig  <- igi(insulin_0 = 10, insulin_t = 80, glucose_0 = 5.3, glucose_t = 8.9)
mat <- matsuda(glucose_0 = 95, insulin_0 = 10,
               glucose_mean = 140, insulin_mean = 60)
disposition_index(secretion = ig, sensitivity = mat)   # oral DI (Matsuda)
```

## HOMA2

HOMA2 has no closed-form equation (the Oxford DTU model's constants were never
published), so SurInsulot bundles the DTU Calculator v2.2.4 reference tables and
interpolates over them. The three input modes are available directly, and
`compute_surrogates()` adds `homa2_ir`, `homa2_b_ins`, and `homa2_s` (plus the
C-peptide variants) automatically.

```{r}
homa2_insulin(glucose = 5.0, insulin = 60)      # glucose mmol/L, insulin pmol/L
homa2_cpeptide(glucose = 5.0, cpeptide = 0.7)   # C-peptide nmol/L
```

Grid-point inputs reproduce the official DTU calculator exactly; values outside
the valid range (glucose 3–25 mmol/L, insulin 20–400 pmol/L, C-peptide
0.2–3.5 nmol/L) return `NA`.

## Optional Yeo-Johnson transform

Many indices are right-skewed, which matters for parametric models. Passing
`transform = TRUE` appends a Yeo-Johnson-normalised `<index>_yj` column for each
index, with lambda fitted per index by maximum likelihood (the fitted values are
stored in `attr(out, "surinsulot_yj_lambda")`). The transform is also available
standalone:

```{r}
x <- c(1, 2, 2, 3, 5, 8, 13, 21, 34, 55)
lam <- yeo_johnson_lambda(x)
round(yeo_johnson(x, lam), 3)
```

When cross-validating, fit lambda on the training fold only, to avoid leakage.
